---
title: 'LLMCellType: Reference-free and cost-effective automated cell type annotation
  with large language models in single-cell RNA-seq analysis'
author:
- Wenpin Hou$^{1,*}$, Zhicheng Ji$^{2,*}$
- $^1$Department of Biostatistics, Columbia University Mailman School of Public Health Health
- $^2$Department of Biostatistics and Bioinformatics, Duke University School of Medicine
- $^*$ corresponding authors
output:
  html_document:
    df_print: paged
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{LLMCellType: reference-free and cost-effective automated cell type annotation with large language models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introductions

Cell type annotation is an essential step in single-cell RNA-seq analysis. However, it is a time-consuming process that often requires expertise in collecting canonical marker genes and manually annotating cell types. Automated cell type annotation methods typically require the acquisition of high-quality reference datasets and the development of additional pipelines. We demonstrated that GPT-4, a highly potent large language model, can automatically and accurately annotate cell types by utilizing marker gene information generated from standard single-cell RNA-seq analysis pipelines in [this manuscript](https://www.biorxiv.org/content/10.1101/2023.04.16.537094v1).
We developed this software, **LLMCellType**, to provide an automated cell type annotation approach using large language models for single-cell RNA-seq analysis. It is provider-agnostic and supports OpenAI GPT, Anthropic Claude, Google Gemini, as well as local models served by Ollama (via the [ellmer](https://ellmer.tidyverse.org) package).


### Installation

LLMCellType can be installed by following [this instruction](https://github.com/pwwang/LLMCellType) on Github.

```{r eval = FALSE}
remotes::install_github("pwwang/LLMCellType")
```

### Set up an API key as an environment variable

LLMCellType connects to the chat API of your provider of choice. To do so, a secret API key is required for OpenAI, Anthropic and Google Gemini. To avoid the risk of exposing the API key or committing the key to browsers, users need to set up the API key as a system environment variable before running LLMCellType. If the API key is provided, cell type annotations are returned. Otherwise, if the API key is not provided, the output from LLMCellType is the prompt itself which users can further use to communicate with the chatbot of their provider. No API key is needed for provider = 'ollama', which runs against a local [Ollama](https://ollama.com) server.

| provider  | API key environment variable | Website to obtain the key |
|-----------|------------------------------|---------------------------|
| openai    | `OPENAI_API_KEY`             | https://platform.openai.com/api-keys |
| anthropic | `ANTHROPIC_API_KEY`          | https://console.anthropic.com/settings/keys |
| gemini    | `GOOGLE_API_KEY` or `GEMINI_API_KEY` | https://aistudio.google.com/apikey |
| ollama    | none (local server)          | - |

For example, for OpenAI you can generate your API key in your OpenAI account webpage: log in to [OpenAI](https://openai.com/). In the pop-up windows, click on "->" next to "API"; next, click on the left-hand-side icon of "API key"; then click on "Create new secret key" to create your key which directs you to the [API key page](https://platform.openai.com/api-keys). Copy the key and paste it on a note for further use. Avoid sharing your API key with others or uploading it to public spaces. Make sure it's not visible in browsers or any client-side scripts. The keys for the other providers are obtained similarly from their respective websites above.

Set up the API key as a system environment variable before running LLMCellType.
```{r,eval=FALSE}
Sys.setenv(OPENAI_API_KEY = 'your_openai_API_key')
# Sys.setenv(ANTHROPIC_API_KEY = 'your_anthropic_API_key')
# Sys.setenv(GOOGLE_API_KEY = 'your_gemini_API_key')
```


### Run LLMCellType

First of all, please load the packages.
```{r}
library(LLMCellType)
```

We demonstrate how to run LLMCellType as follows.  The main function is **`llmcelltype()`**. It can annotate cell types by LLM chat models in a Seurat pipeline or with a custom gene list. If **`llmcelltype()`** is used in a Seurat pipeline, Seurat **`FindAllMarkers()`** function needs to be run first and the differential gene table generated by Seurat will serve as the input. If the input is a custom list of genes, one cell type is identified for each element in the list.

Among the input arguments, **`input`** can either be the differential gene table returned by Seurat **`FindAllMarkers()`** function, or a list of genes.
**`tissuename`** (optional) is a tissue name.
**`provider`** is one of 'openai' (default), 'anthropic', 'gemini' or 'ollama'.
**`model`** is a model name accepted by the provider. If `NULL` (default), the provider's default model is used. provider = 'ollama' always requires an explicit model.
**`topgenenumber`** is the number of top differential genes to be used when the input is Seurat differential genes. The output is a vector of cell types.


#### Example 1: Seurat object as input

LLMCellType integrates seamlessly with the Seurat pipeline. It can take a Seurat object as input, if the Seurat object has marker genes information. Specifially, this can be achieved after running the Seurat function `FindAllMarkers()`. Here follows an example.

Load the Seurat package.

```{r}
library(Seurat, quietly = TRUE)
```



In the below example, we are going to use a Seurat object called 'pbmc_small' provided by the Seurat package. In real applications, a Seurat project obtained after running the standard Seurat pipeline should be prepared. The Seurat project should have cell clustering available. Use FindAllMarkers() function to generate the differential gene table if you haven't done so:

```{r}
data("pbmc_small")
suppressWarnings({
  all.markers <- FindAllMarkers(object = pbmc_small)
})

```

Perform cell type annotation by an LLM using the llmcelltype() function. Here you can optionally provide the actual name of the tissue for your dataset.
```{r}
res <- llmcelltype(all.markers,
            tissuename = 'human PBMC'
)
```

It is always recommended to check the results returned by the model in case of AI hallucination, before going to down-stream analysis.
```{r}
res
```

If the results make sense, we can assign the cell type annotations back to the Seurat object and visualize the cell type annotations on the UMAP:

```{r,  fig.height=4}
pbmc_small@meta.data$celltype <- as.factor(res[as.character(Idents(pbmc_small))])
DimPlot(pbmc_small,group.by='celltype')
```

If the results need to be fine-tuned, it is easy to reassign cell type annotations for some clusters. For example, to change the cell type annotation for cluster 0:

```{r}
res[1] <- 'Classical monocytes'
pbmc_small@meta.data$celltype <- res[as.character(Idents(pbmc_small))]
```

If you prefer not to link to the API of your provider or do not have an API key, you can set the key environment variable to `''` (e.g. `Sys.setenv(OPENAI_API_KEY = '')`). In this case, the llmcelltype() function will print the prompt directly, which can be copied and pasted into the online chat interface of your provider to obtain cell type annotations.
```{r}
Sys.setenv(OPENAI_API_KEY = '')
data("pbmc_small")
suppressWarnings({
  all.markers <- FindAllMarkers(object = pbmc_small)
})
res <- llmcelltype(all.markers,
            tissuename = 'human PBMC'
)
cat(res)
```


####  Example 2: use a list of genes as input

Set up the API key of your provider as a system environment variable before running LLMCellType.
```{r,eval=FALSE}
Sys.setenv(OPENAI_API_KEY = 'your_openai_API_key')
```

If we provide a list of two gene vectors: the first vector contains *CD4* and *CD3D*, and the second vector contains *CD14*, then we can call the function in this way:
```{r}
res <- llmcelltype(
  input = list(cluster1 = c('CD4, CD3D'), cluster2 = 'CD14'),
  tissuename = 'human PBMC'
)
res
```

#### Example 3: other providers

The same calls work with any supported provider by setting `provider` and the corresponding API key environment variable. No API key is required for provider = 'ollama' (make sure the Ollama server is running locally and the model is pulled first).

```{r eval = FALSE}
# Anthropic Claude
res <- llmcelltype(all.markers, tissuename = 'human PBMC', provider = 'anthropic')

# Google Gemini
res <- llmcelltype(all.markers, tissuename = 'human PBMC', provider = 'gemini')

# Local Ollama server
res <- llmcelltype(all.markers, tissuename = 'human PBMC',
                   provider = 'ollama', model = 'llama3.1')
```

Any OpenAI-compatible endpoint (OpenRouter, DeepSeek, local servers, ...) can be used with `provider = 'openai'` plus a custom `base_url`, e.g.:

```{r eval = FALSE}
res <- llmcelltype(all.markers, tissuename = 'human PBMC',
                   base_url = 'https://api.deepseek.com/v1',
                   model = 'deepseek-chat')
```

For backward compatibility, `gptcelltype()` is retained as an alias of `llmcelltype(provider = 'openai')`.

#### Session Info
```{r}
sessionInfo()
```
